
%%%%%%%%%%
% Lang Ref Man
%%%%%%%%%%
\chapter{Language Reference Manual}\label{refman}

\section{Preamble}
Herein is the language reference manual for the BALL programming
language. As a language, BALL is designed for user simplicity and ease
of programming. Thus, users can create simulations and manipulate
statistics with ease by using the constructs presented below.

\subsection{Purpose of the Language}
The main purpose of BALL is to allow users to easily construct
simulations of baseball games. BALL defines language constructs that
help the user acheive this goal this goal. The design of the BALL
language also has in mind other related objectives, such as allowing
statistical data to be easily viewed and compared. To do this, BALL
encapsulates many low-level tasks that are present in more general
programming languages.

\subsection{BALL Compilation Process}\label{ref_compile}
BALL is a compiled language. Source code is stored in files ending
with the \texttt{.ball} extension (for example: \texttt{hello.ball}),
and is provided to the BALL compiler. Valid BALL source files must end
with a newline. The compiler will then automatically generate the
low-level code that is syntactically equivalent to the original source
and execute it. Our reference implementation of a BALL compiler uses
Java as a target language.

BALL is an imperative programming language. It is similar in structure
to Java, though numerous changes and innovations have been made to
help the user quickly learn and create BALL programs.

\subsection{Abstract Features}
\subsubsection{Types}\label{types}
BALL is a typed language, specifically with five types,
\texttt{number}, \texttt{string}, \texttt{list}, \texttt{team}, and
\texttt{player}. The \texttt{number} and \texttt{string} types are
simple constants. This is further explained in the section
\ref{keywords} below. The type \texttt{list} is a collection of
objects, which are instances of types. \texttt{list} is therefore
typed itself (see \ref{of}). Thus it is possible to have a
\texttt{list} of \texttt{team} objects, \texttt{player} objects,
\texttt{string}s, \texttt{number}s, or other \texttt{list}
objects. \texttt{list} objects are homogenous, meaning they cannot
contain objects of multiple types. The \texttt{player} type is an
object that has attributes of \texttt{string} and \texttt{number}
types. These attributes include the name of the player, whether the
player is a pitcher or not, and the statistics of that player. The
\texttt{team} type is an object that has attributes of
\texttt{string}, \texttt{number}, and \texttt{list} types. These
attributes include the name of the team, the team's wins and losses,
and the list of every player on the team.


\input{refman/lexical.tex}

\input{refman/expr.tex}

\input{refman/decl.tex}

\input{refman/stmt.tex}

\input{refman/builtin.tex}

\input{refman/csv_spec.tex}

\pagebreak

\section{Complete Grammar}\label{grammar}

\codelisting{FullGrammar}{Full Grammar}{../docs/ball_full.y}

