\chapter{Project Plan}

\section{Development Process}\label{Process}

In developing the BALL language, we followed the suggestion presented
by Prof. Aho during class: make a working prototype, and incrementally
add enhancements. During the early parts of the development, more
attention is focused on adding new features than testing, however each
team member must create his own test case with the changes he made.

Whenever a part of the language is finished, it is tested and reviewed
by the team. Feedback is then relayed to the team member responsible
for the subsystem (usually the one who merged the latest version to
trunk, as many modules are jointly developed by the team). There is
less of a focus on the traditional team roles, rather each team member
focuses on developing the compiler while also doing their job in team
administration.

\pagebreak
\section{Style Sheet}\label{Style}

\begin{singlespacing}
\begin{verbatim}
/*
 * COMS W4119 PROGRAMMING LANGUAGES AND TRANSLATORS FALL 2009
 * Team llamamelon - BALL language
 * StyleSheet.java - style sheet for Java classes
 */

package style;

// package declaration comes after header, separated by one line
/*
 * header contains course name, team name, language name, file name, and a short
 * description of what the file is.
 */

// imports come right after, separated by at least 1 empty line
import java.util.Arrays;

import java.io.*;
// imports can have empty lines separating them, especially when importing
// different trees

import javabackend.SimFunction;
import javabackend.TeamObj;

// single line comments can use // or /* (for more important comments)
// two-line comments written like this are also permitted. For 3 or more, use /*

/*
 * Very important single-line comments are made like this
 */

/*
 * Multi-line comments also use this format. The maximum width for each line in
 * a multi-line comment is 80 characters from the start of the line (not the
 * start of the comment, which is after the space after a '*'.
 * 
 * 1.  this is how to write lists with numbers.
 * 2.  When a list goes over a line, the next line starts under the start of the
 *     first line of the list.
 *     ...
 * 10. If the numbering goes over 10, entries 1 to 9 need to have 1 more space.
 *     This way all list text starts at the same place. Having a list of more 
 *     than 10 members is discouraged.
 */

/**
 * Comments before a class uses Javadoc style. It is REQUIRED, even if just one
 * sentence. Give a brief description of what role it has in the compiler. 
 * 
 * Class names are capitalized on each word. 
 * - playerObj - WRONG 
 * - PlayerObj - RIGHT
 */
public class StyleSheet extends StyleSuper implements StyleInterface {

    // comments follow the indentation of the non-commented statements
    
    // public constructor comes first.
    /**
     * Sets use to default value. 
     */
    public StyleSheet() {
        this.use = 1001;
    }

    /*
     * Comments describing a method (highly recommended for public methods, even
     * for just one line). Some details:
     *
     * 1. comments for private methods are recommended.
     * 2. Javadoc style is optional. Can both be /* or /** or even //
     * 3. comments for methods that override a superclass method / implement an
     *    interface method are highly recommended it's not already documented in
     *    the superclass / interface. Otherwise, it's optional.
     * 5. Methods that override builtin classes' methods like toString() or
     *    hashCode() don't need to be commented.
     * 
     * Method names
     * 1. first letter of name NOT capitalized. However, words after that are
     *    captialized.
     * 2. Order of methods:
     *    a. public
     *    b. protected
     *    c. <no scope>
     *    d. private
     *    e. public static
     *    ...
     *    h. private static
     *    the names themselves can go in any order.
     * 
     * 1. Annotations like @Override are OPTIONAL. They go after the comment.
     * 2. Public methods cannot have underscores.
     */
    public int sampleFun(int arg1, int arg2, Object arg3) {
        // no space here ^ between function name and opening parentheses.
        // single space between arguments, and btwn close paren and open brace

        int p = 0, q; // space after each variable name
        q = 10;
        
        // note the spaces on for loops
        for (int i = 0; i < 10, i++) {
            // inside of for loops are also indented
            p++;
            q += 3;
            if (q >= 20) // single-line ifs like this are permitted
                System.out.println("bleh");
        }
        
        /*
         * Inner variables don't have to be capitalized properly.
         * 
         * Anonymous classes have a space after parentheses and opening curly
         * brace. The inside is indented.
         */
        SimFunction mysim = new SimFunction() {
            public TeamObj doSim(TeamObj team1, TeamObj team2) {
                return team1;
            }
        };

        return 1037; // parentheses for return exprs aren't necessary
    }
    
    /**
     * Since toString() is already in object, this is technically unnecessary
     */
    @Override
    public String toString() {
        return "style sheet";
    }
    
    /*
     * Order of fields is just like order of methods. "final" goes after
     * non-final but inside the same bracket above. So, "protected final" goes
     * after "public final" or "protected" but before "private".
     * 
     * protected x
     * public static final y
     */
    public int use;
    
    public final String bleh = "bleh";
    
    private float[] arrayExample;
    // variables/fields are named like methods

}
\end{verbatim}
\end{singlespacing}

\pagebreak

\section{Timeline \& Logs}\label{TimelineLogs}
\subsection{Timeline of BALL Development}

\begin{center}
\begin{longtable}{|c|p{10cm}|}
\hline
Time & Event\\
\hline
9/18/09, 11:31a & Team Llamamelon Born.\\
9/21/09, 5:15p & First Team Llamamelon Meeting; Team Llamamelon
named. Sam pioneers the language purpose: the simulation of baseball
games.\\
9/21/09, 8:02p & Team Llamamelon announced to PLT Teaching Staff.\\
10/2/09, 4:42p & Alfred V. Aho tentatively approves BALL programming
language.\\
10/5/09, 1:04a & Buzzwords for initial Whitepaper finalized.\\
10/7/07, 3:17p & Final Whitepaper completed.\\
10/23/09, 2:52p & Sample programs designed, BALL envisioned to be a
scripting language à la AWK.\\
10/27/09, 4:14p & Google Code Group Created. Jordan initializes
Subversion repository and trains team for first 'commit.'\\
10/27/09, 5:30p & First major Team Llamamelon Meeting; language
roadmap created.\\
10/29/09, 10:30a & First meeting with Sinan; BALL changes from
scripted language to compiled language, based on Java.\\
10/30/09, 1:27p & Nathan creates first example program; the new
language roadmap is born.\\
10/31/09, 10:42p & Cipta begins creating the grammar, laying the
groundwork for BALL.\\
11/3/09, 1:05p & Language Reference Manual and Tutorial begun.\\
11/3/09, 4:43p & Language Grammar completed, Daniel creates and
finalizes complex simulation to show off the powerful grammar.\\
11/4/09, 10:26a & Final Language Reference Manual and Tutorial
completed.\\
11/15/09, 3:00p & First semi-weekly meeting. Full grammar implemented
using JFlex/BYACC-J. Lexer created a DFA!\\
11/18/09, 8:00p & Second semi-weekly meeting. Play.ball designed,
intermediate code generation begins.\\
11/19/09, 11:00a & Second meeting with Sinan; Language Reference
Manual and Tutorial approved; BALL is in full motion!\\
11/27/09, 12:43a & Play.Ball compiles and runs: ``Play Ball!'' [Hello
  world equivalent.]\\
12/3/09, 3:45p & Team Loader created; teams can now load into the
PlayerObj and TeamObj constructs.\\
12/5/09, 11:46a & Major all-day meeting; BALL presentation created and
beautified.\\
12/7/09, 4:30p & BALL presented to PLT. There was much rejoicing.\\
12/10/09, 4:37a & SymbolTable embraced as major component of BALL.\\
12/10/09, 3:02a & Simfunc.ball compiles and runs: simfunctions and
activate work.\\
12/12/09, 10:00a & Major all-day meeting; Much of language completed
during this time.\\
12/12/09, 10:48a & TeamStat and PlayerStat methods added: stats work
throughout BALL.\\
12/12/09, 7:59p & Lists become typed, major change in the language.\\
12/12/09, 8:55p & Logical, Boolean, Comparison expressions completed.\\
12/15/09, 12:56a & \texttt{where}, \texttt{from}, and \texttt{any} keywords working.\\
12/16/09, 2:02a & \texttt{'s} keyword works for all attributes, stats, and
lists inherent in team and player objects.\\
12/16/09, 6:08p & Unary operators completed.\\
12/17/09, 6:27p & Realized the class number is W4115, not
W4119. [Seriously.]\\
12/20/09, 10:30a & Final Major all-day meeting: everything completed.\\
12/20/09, 8:04p & Everything in the language works. Final presentation
compilation begun.\\
Now & Final presentation completed.\\

\hline
\caption{Token Example}
\label{token example}
\end{longtable}
\end{center}

\subsection{Project Log}

\input{projlog.tex}
